---
title: "PCA Analysis - TB and 1000 Genomes"
#author: "Ping Zhang"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
    code_folding: show
params:
  PCA_1KG_EIGENVEC: "/well/jknight/users/kwz374/EQTL_OXFORD/Analysis_freeze.2024/Imputation/Michigan/PCA.with.1KG.data/plink.eigenvec"
  PCA_1KG_EIGENVAL: "/well/jknight/users/kwz374/EQTL_OXFORD/Analysis_freeze.2024/Imputation/Michigan/PCA.with.1KG.data/plink.eigenval"
  PCA_TB_EIGENVEC: "/well/jknight/users/kwz374/EQTL_OXFORD/Analysis_freeze.2024/LMM.analysis/TB.ALL.eigenvec"
  PCA_TB_EIGENVAL: "/well/jknight/users/kwz374/EQTL_OXFORD/Analysis_freeze.2024/LMM.analysis/TB.ALL.eigenval"
  SAMPLE_SHEET: "/well/jknight/users/kwz374/EQTL_OXFORD/Analysis_freeze.2024/JupyterLab/temp/1_metadata_2024_swap.corrected..xlsx"
  SHEET_NUM: 2
  RESULTS_DIR: "/well/jknight/users/kwz374/EQTL_OXFORD/Analysis_freeze.2024/JupyterLab/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 6,
  fig.height = 6
)

# Create results directory
dir.create(params$RESULTS_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Introduction
This report performs Principal Component Analysis (PCA) on TB samples combined with 1000 Genomes data to visualize population structure and identify significant principal components.

## Load Libraries

```{r libraries}
library(openxlsx)
library(ggplot2)
library(ggrepel)
library(scatterplot3d)
library(data.table)
library(reshape)
```

## Part 1: PCA with 1000 Genomes Reference

### Load PCA Data and Metadata

```{r load_pca_1kg}
# Read in the eigenvectors produced in PLINK
pca <- read.table(params$PCA_1KG_EIGENVEC, 
                  header = FALSE, skip = 0, sep = ' ')
rownames(pca) <- pca[,2]
pca <- pca[,3:ncol(pca)]
colnames(pca) <- paste('PC', c(1:20), sep = '')

# Read in the PED data
PED <- read.xlsx(params$SAMPLE_SHEET, sheet = params$SHEET_NUM)
PED <- PED[which(PED$Individual_ID %in% rownames(pca)), ]
PED <- PED[match(rownames(pca), PED$Individual_ID),]
cat("Sample IDs match:", all(PED$Individual_ID == rownames(pca)), "\n")
```

### Prepare Data for Visualization

```{r prepare_data_1kg}
# Merge PCA results with metadata
pca$sample.name = rownames(pca)
plot <- merge(pca, PED, by.x = "sample.name", by.y = "Individual_ID")

# Create super population categories
plot$super.population <- ifelse(plot$Population %in% c("ACB","ASW","ESN","GWD","LWK","MSL","YRI"), "AFR",
                                ifelse(plot$Population %in% c("CLM","MXL","PEL","PUR"), "AMR",
                                       ifelse(plot$Population %in% c("CDX","CHB","CHS","JPT","KHV"), "EAS",
                                              ifelse(plot$Population %in% c("CEU","FIN","GBR","IBS","TSI"), "EUR",
                                                     ifelse(plot$Population %in% c("BEB","GIH","ITU","PJL","STU"), "SAS", "HIRV TB")))))

cat("Super population distribution:\n")
print(table(plot$super.population))

# Calculate percentage variance explained
eigenval <- scan(params$PCA_1KG_EIGENVAL)
percentVar <- data.frame(PC = 1:20, pve = sprintf("%.1f", eigenval/sum(eigenval)*100))

# Set factor levels
plot$super.population = factor(plot$super.population, 
                               levels = c("AFR", "AMR", "EAS", "EUR", "SAS", "HIRV TB"))
```

### 2D PCA visualisation

```{r pca_2d_1kg}
# PC1 vs PC2
ggplot(plot, aes(x = PC1, y = PC2, color = super.population)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: TB Samples with 1000 Genomes Reference",
       x = paste0("PC1 (", percentVar$pve[1], "%)"),
       y = paste0("PC2 (", percentVar$pve[2], "%)"),
       color = "Super Population") +
  theme_minimal() +
  theme(legend.position = "bottom")

# PC3 vs PC4
ggplot(plot, aes(x = PC3, y = PC4, color = super.population)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "PCA: PC3 vs PC4",
       x = paste0("PC3 (", percentVar$pve[3], "%)"),
       y = paste0("PC4 (", percentVar$pve[4], "%)"),
       color = "Super Population") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### 3D PCA visualisation

```{r pca_3d_1kg}
# 3D PCA for HIRV TB samples only
plot2 = plot[plot$super.population %in% "HIRV TB",]
colors <- c("#999999", "#E69F00", "#56B4E9", "red", "purple")
colors <- colors[as.numeric(as.factor(plot2$Population))]

s3d <- scatterplot3d(plot2$PC1, plot2$PC2, plot2$PC3,
                    xlab = "PC1", ylab = "PC2", zlab = "PC3", 
                    pch = 16, color = colors, main = "3D PCA: HIRV TB Samples")
legend("right", legend = levels(as.factor(plot2$Population)),
       col = c("#999999", "#E69F00", "#56B4E9", "red", "purple"), pch = 16)
```

## Part 2: PCA with TB samples only

### Load TB-only PCA Data

```{r load_pca_tb}
# Read in eigenvectors for TB samples only
pca_tb <- read.table(params$PCA_TB_EIGENVEC, 
                     header = FALSE, skip = 0, sep = ' ')
rownames(pca_tb) <- pca_tb[,2]
pca_tb <- pca_tb[,3:ncol(pca_tb)]
colnames(pca_tb) <- paste('PC', c(1:20), sep = '')

# Merge with metadata
pca_tb$sample.name = rownames(pca_tb)
plot_tb <- merge(pca_tb, PED, by.x = "sample.name", by.y = "Individual_ID")

cat("Population distribution in TB samples:\n")
print(table(plot_tb$Population))

# Calculate percentage variance explained for TB-only data
eigenval_tb <- scan(params$PCA_TB_EIGENVAL)
percentVar_tb <- data.frame(PC = 1:20, pve = sprintf("%.1f", eigenval_tb/sum(eigenval_tb)*100))
```

### TB-only PCA visualisation

```{r pca_tb_only}
# 2D PCA for TB samples
ggplot(plot_tb, aes(x = PC1, y = PC2, color = Population)) +
  geom_point(alpha = 0.7, size = 3) +
  labs(title = "PCA: TB Samples Only",
       x = paste0("PC1 (", percentVar_tb$pve[1], "%)"),
       y = paste0("PC2 (", percentVar_tb$pve[2], "%)")) +
  theme_minimal() +
  theme(legend.position = "bottom")

# 3D PCA for TB samples
colors_tb <- c("#999999", "#E69F00", "#56B4E9", "red", "purple")
colors_tb <- colors_tb[as.numeric(as.factor(plot_tb$Population))]

s3d_tb <- scatterplot3d(plot_tb$PC1, plot_tb$PC2, plot_tb$PC3,
                       xlab = "PC1", ylab = "PC2", zlab = "PC3", 
                       pch = 16, color = colors_tb, 
                       main = "3D PCA: All TB Samples")
legend("right", legend = levels(as.factor(plot_tb$Population)),
       col = c("#999999", "#E69F00", "#56B4E9", "red", "purple"), pch = 16)
```

## Part 3: determine significant PCs associated with population structure
```{r pc_association}
# Test each PC for association with population
dat2 = list()
t = colnames(plot_tb[,2:21]) # 20 PCs

for (i in t) {
  formula <- as.formula(paste(i, "~ Population"))
  fit = lm(formula, data = plot_tb)
  
  dat2[[i]] = data.frame(
    beta.AMR = coef(fit)[2],
    beta.EAS = coef(fit)[3],
    beta.EUR = coef(fit)[4],
    beta.SAS = coef(fit)[5],
    p.valueAMR = coef(summary(fit))[2,4],
    p.valueEAS = coef(summary(fit))[3,4],
    p.valueEUR = coef(summary(fit))[4,4],
    p.valueASA = coef(summary(fit))[5,4],
    PC = gsub("PC", "", i)
  )
}

dat2 = as.data.frame(do.call(rbind, dat2))

# Adjust p-values for multiple testing
dat2$adjp.AMR = p.adjust(dat2$p.valueAMR, method = "fdr")
dat2$adjp.EAS = p.adjust(dat2$p.valueEAS, method = "fdr")
dat2$adjp.EUR = p.adjust(dat2$p.valueEUR, method = "fdr")
dat2$adjp.ASA = p.adjust(dat2$p.valueASA, method = "fdr")

cat("Significant PC associations found:\n")
print(dat2[dat2$adjp.AMR < 0.05 | dat2$adjp.EAS < 0.05 | 
           dat2$adjp.EUR < 0.05 | dat2$adjp.ASA < 0.05, ])
```

### Visualise significant PCs
```{r visualize_significant_pcs}
# Prepare data for visualization
df = dat2[,c("adjp.AMR", "adjp.EAS", "adjp.EUR", "adjp.ASA", "PC")]
df = melt(df, id = "PC")

# Create bar plot of adjusted p-values
ggplot(df, aes(reorder(PC, as.numeric(PC)), -log10(value), fill = variable)) +
  geom_bar(stat = "identity", position = position_dodge()) +  
  theme_bw() + 
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Principal Component", 
       y = "-Log10(Adjusted P-value)",
       title = "Association of PCs with Population Structure",
       fill = "Population") +
  geom_hline(yintercept = -log10(0.05), linetype = "dotted", color = "red")
```

## Summary
```{r summary}
cat("### Analysis Summary\n")
cat("- Total samples in 1KG + TB analysis:", nrow(plot), "\n")
cat("- Total samples in TB-only analysis:", nrow(plot_tb), "\n")
cat("- Number of significant PC-population associations:", 
sum(dat2$adjp.EAS < 0.05 | dat2$adjp.EUR < 0.05 | dat2$adjp.ASA < 0.05), "\n")
#cat("- PCs with ≥2 adj-p < 0.05:",
#    sum(rowSums(dat2[, c("adjp.AMR","adjp.EAS","adjp.EUR","adjp.ASA")] < 0.05, na.rm = TRUE) >= 2), "\n")

```

## Session Information

```{r session_info}
sessionInfo()
```